---
title: "Module 3 — Data wrangling"
format: html
editor: source
execute:
  warning: false
  message: false
---

> *Your mode for this exercise.* Group A: **agentic** — describe each step to Pi and let it write the code. Group B: **manual** — write each step yourself. A chat LLM may answer conceptual questions but should not edit this notebook or run commands. Exercise 2 flips the modes.

```{r setup}
library(tidyverse)
library(here)

table_02 <- read_csv(here("data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))
```

## Part 1 — Predict (no code)

Take a look at `table_02` (run the cell above first). Answer in plain words below — no code yet.

```{r}
table_02
```

**1. How many rows? What does one row represent?**

The table has 132 rows. Each row represents details for each participant.

**2. After `table_02 |> filter(time_point == "baseline")`, how many rows? What does each row represent now?**

This picks rows that have baseline for the timepoint. 

**3. After `table_02 |> count(arm)`, how many rows? What does each row represent now?**

 This counts how many rows have treatment arm.

## Part 2 — Build the pipeline

Each step adds one verb on top of the previous pipeline. Run after each step.

```{r}
table_02
```

```{r}
table_02 |> filter (time_point == "baseline")
```

```{r}
table_02 |> select (pid, arm, ph,nugent_score, crp_blood) 
```

```{r}
table_02 |> mutate (high_ph = if_else(ph>=4.5, "yes", "no"))
```

```{r}
table_02 |> arrange(desc(ph))
```

```{r}
# step 6: assign the final pipeline to early_visit, then print it
early_visit <- NULL  # replace NULL with the pipeline
early_visit
```

```{r}
# step 7: separately, table_02 |> count(arm)
```

## Part 3 — Reach (optional)

Add a `nugent_category` column with `case_when()`:

- 0–3 → "normal"
- 4–6 → "intermediate"
- 7–10 → "BV"

Then run `count(arm, nugent_category)`.

```{r}
# reach: mutate(nugent_category = case_when(...)) then count(arm, nugent_category)
```
